Beyond May: Complexity-stability relationships \\ in disordered dynamical systems

### 2023-05-03 papers/abstract.tex
captions and refs abbreviated

Robert May famously used random matrix theory to predict that large, complex systems cannot admit stable fixed points. 
However, this seemingly general conclusion is not supported by empirical observation: from cells to biomes, biological systems are large, complex---and, by and large, stable.
In this paper, we revisit May's argument in the light of recent developments in both ecology and random matrix theory. 
Using a non-linear generalization of the Lotka-Volterra model, we show that there are in fact two kinds of complexity-stability relationships in disordered dynamical systems: 
if self-interactions grow faster with density than cross-interactions, complexity is destabilizing; but if cross-interactions grow faster than self-interactions, complexity is stabilizing.
Our result shows that May's principle that "complexity begets instability" is not a general property of complex systems; instead, it is a property of a class of weakly cross-regulated complex systems.

>>>>

Robert May famously used random matrix theory to predict that large, complex systems cannot admit stable fixed points. 
Unfortunately, this general conclusion is not supported by empirical observation: from cells to biomes, biological systems are large, complex---and, by and large, stable.
In this paper, we revisit May's argument in the light of recent developments in both ecology and random matrix theory. 
Using a non-linear generalization of the Lotka-Volterra model, we show that there are in fact two kinds of complexity-stability relationships in disordered dynamical systems: 
if self-interactions grow faster with density than cross-interactions, complexity is destabilizing; but if cross-interactions grow faster than self-interactions, complexity is stabilizing.
Our result shows that May's principle that "complexity begets instability" is not a general property of complex systems; instead, it is a property of a subclass of weakly cross-regulated disordered systems.

### 2023-05-03 papers/beyond-may.tex
captions and refs abbreviated

@@

\author{Onofrio Mazzarisi}
\affiliation{Department of Ecology and Evolutionary Biology, University of Kansas, Haworth Hall, 
1200 Sunnyside Avenue
Lawrence, KS 66045, USA}

@@

\paragraph*{\bf Introduction.}
\input{intro.tex}

\paragraph*{\bf Model.}
\input{model.tex}

\paragraph*{\bf Homogeneous interactions.}
\input{homogeneous.tex}

\paragraph*{\bf Random interactions: stability condition.}
\input{stability.tex}

\paragraph*{\bf Random interactions: equilibrium distribution.}
\input{equilibrium.tex}

\paragraph*{\bf Conclusion.}
\input{conclusion.tex}

@@

\begin{acknowledgments}
We thank Ian Hatton, Matthieu Barbier and Ada Altieri for many productive discussions on ecological power laws and the diversity-stability debate. 
\end{acknowledgments}

>>>>

\captionsetup{justification=Justified}

@@

\author{Onofrio Mazzarisi}
\affiliation{Department of Ecology and Evolutionary Biology, University of Kansas,
Higuchi Hall, 2101 Constant Avenue, Lawrence, KS 66047, USA}

@@

%Introduction
\paragraph*{}
\input{intro.tex}

%Model
\paragraph*{}
\input{model.tex}

%Homogeneous interactions
\paragraph*{}
\input{homogeneous.tex}

%Random interactions: stability condition
\paragraph*{}
\input{stability.tex}

%Random interactions: equilibrium distribution
\paragraph*{}
\input{equilibrium.tex}

%Conclusion
\paragraph*{}
\input{conclusion.tex}

@@

\begin{acknowledgments}
We thank Ian Hatton, Matthieu Barbier and Ada Altieri for productive discussions on ecological power laws and the diversity-stability debate. 
\end{acknowledgments}

### 2023-05-03 papers/model.tex
captions and refs abbreviated

This rather general setting has been used to study universality in network dynamics \cite{Barzel2013} and to construct minimal models for complex dynamics \cite{Barzel2015}, with applications ranging from biological~\cite{Alon2006,Karlebach2008,Skinner2012} 
to social~\cite{Pastor-Satorras2001,Hufnagel2004,Dodds2005} systems.

>>>>

This rather general setting has been used to study universality in network dynamics \cite{Barzel2013} and to construct minimal models for complex dynamics \cite{Barzel2015}, with applications ranging from biological~\cite{Alon2006,Karlebach2008} 
to social~\cite{Pastor-Satorras2001,Hufnagel2004,Dodds2005} systems.
